---
title: "Getting Started with spawnBayes: A Clemens Creek Sockeye Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spawnBayes: A Clemens Creek Sockeye Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, message = FALSE, warning = FALSE}
library(spawnBayes)
library(ggplot2)
library(dplyr)
library(lubridate)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

spawnBayes estimates salmon spawner abundance and run timing from repeated
live-count surveys. Unlike traditional area-under-the-curve approaches,
spawnBayes estimates abundance, run timing, and their associated uncertainty
while allowing residence time to vary among years.

The underlying methodology is described in Thompson et al. (2026).

This vignette demonstrates a complete analysis using sockeye salmon survey
data from Clemens Creek, BC, Canada. These data were provided by Diana
McHugh, South Coast Stock Assessment, Fisheries and Oceans Canada.

# Workflow

A typical spawnBayes analysis consists of:

1. Preparing survey observations.
2. Validating the data structure.
3. Fitting the Bayesian model.
4. Evaluating diagnostics.
5. Estimating abundance and run timing.

# Example data
Load in the example dataset and inspect it. The model requires a date column and a spawner_counts column. Optional observer_efficiency and coverage columns can be supplied to adjust observed counts.

In this case, dates are provided as character strings, and observer efficiency and coverage have been provided as percentages. These will be converted into the proper format in the data validation step.


```{r load-data}
data(clemens_sockeye)

summary(clemens_sockeye)
```

# Validate data
The first step is to validate and standardize the input data. This converts the dates into their proper format and converts observer efficiency and coverage to proportions if they were provided as percentages. Validation may generate warnings if missing correction factors are supplied or if some years show potential evidence of multiple run peaks.
```{r validate-data}
clemens_sockeye <- validate_data(clemens_sockeye)

summary(clemens_sockeye)
```

# Survey observations
Now plot the raw survey counts by date in each year.
```{r plot-data, fig.width=8, fig.height=5}
ggplot(clemens_sockeye |> mutate(year = year(date), day = yday(date)),
  aes(x = day, y = spawner_counts, fill = observer_efficiency))+
  geom_point(pch = 21, size = 2) +
  facet_wrap(~year, scale = "free_y")+
  theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(
      labels = scales::label_number(
        scale_cut = scales::cut_short_scale()
      ),
      name = "Spawners observed"
    ) +
    scale_x_continuous(
      labels = function(x)
        format(
          as.Date(x, origin = "2023-12-31"),
          "%b"
        ),
      name = NULL
    )+
  scale_fill_viridis_c(option = "G", direction = -1)
```

# Fit the model
spawnBayes estimates annual run timing and abundance jointly across years.
Information is shared among years through a hierarchical model, improving
estimation in years with sparse survey data.

The settings used in this vignette are intended to keep computation time
reasonable. For applied analyses, users should always examine the
diagnostic output and increase the number of iterations when recommended. 
Even so, the model may require several minutes to fit.
```{r fit-model, cache=TRUE, message=FALSE, warning=FALSE}
fit <- fit_spawner(clemens_sockeye, iter_warmup = 500, iter_sampling = 500, assumed_residence = 15.5, adapt_delta = 0.99, refresh = 0)
```

# Model diagnostics

The `diagnostics()` function summarizes several indicators of MCMC sampling performance and model convergence. 
```{r diagnostics}
diagnostics(fit)
```
In this example, all diagnostic checks were acceptable. No divergent
transitions were detected, no maximum tree-depth warnings were encountered,
and the largest Rhat value was close to 1. Effective sample sizes for both
bulk and tail regions of the posterior distribution were sufficient for
reliable inference.

When fitting new datasets, users should pay particular attention to:

* Divergences: should be zero whenever possible.
* Tree depth hits: frequent hits may indicate that model settings should be adjusted.
* Rhat: values close to 1 indicate convergence, while values substantially
  greater than 1 suggest poor mixing among chains.
* Effective sample size (ESS): larger values indicate better estimation of
  posterior summaries.

The recommendation column provides a summary of whether additional model
tuning or longer runs are required.

# Estimated run
The fitted curves represent the estimated number of spawners present
through time. Shaded regions show posterior 95% credible intervals.
Observed survey counts are shown as points. The open circles show the uncorrected survey counts. The filled circles are corrected for observer efficiency and survey coverage. 

```{r run-curves, fig.width=8, fig.height=7}
plot_run_curve(fit, CI = 95)
```


# Abundance estimates
Estimated annual abundance is obtained by integrating the estimated arrival and exit processes over the spawning season.

```{r abundance}
abund <- abundance(fit)
abund
```

The abundance plot shows the median (point) as well as the 66 and 95% posterior credible intervals. The red x shows the estimate based on trapezoidal area under the curve, with an assumed 15.5 day residence (survey life).
```{r abund-plot, fig.width=6, fig.height=4}
plot_abundance(fit)
```
# Timing estimates

Run timing summaries describe the estimated distribution of spawner arrival dates, residence, and exit lag in each year.

```{r timing-plot, fig.width=8, fig.height=8}
plot_timing(fit)
```

# Conclusions

spawnBayes provides a framework for estimating salmon spawner abundance
and run timing from repeated live-count surveys while accounting for
uncertainty in both the observation and process components of the model.

# Citation

The methods implemented in `spawnBayes` are described in:

Thompson, P. L., Akenhead, S. A., and Louie, C. (2026).
*Estimating spawning salmon abundance from visual surveys:
a hierarchical model of arrival and exit dynamics*.
Canadian Journal of Fisheries and Aquatic Sciences, 83, 1-13.
https://doi.org/10.1139/cjfas-2026-0029

Users are encouraged to cite both the package and the accompanying
manuscript when using spawnBayes in scientific publications.