---
title: "Getting Started with spawnBayes: A Clemens Creek Sockeye Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spawnBayes: A Clemens Creek Sockeye Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, message = FALSE, warning = FALSE}
library(spawnBayes)
library(ggplot2)
library(dplyr)
library(lubridate)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

spawnBayes estimates salmon spawner abundance and run timing from repeated
live-count surveys. Unlike traditional area-under-the-curve approaches,
spawnBayes estimates abundance, run timing, and their associated uncertainty
while allowing residence time to vary among years.

The underlying methodology is described in Thompson et al. (2026).

Additional features introduced after Thompson et al. (2026) include support for survey-error groups, allowing separate observation-dispersion parameters to be estimated for different survey types or quality classes, and an optional robust observation model that reduces the influence of anomalous survey counts.

This vignette demonstrates a complete analysis using sockeye salmon survey
data from Clemens Creek, BC, Canada. These data were provided by Diana
McHugh, South Coast Stock Assessment, Fisheries and Oceans Canada.

# Workflow

A typical spawnBayes analysis consists of:

1. Preparing survey observations.
2. Validating the data structure.
3. Fitting the Bayesian model.
4. Evaluating diagnostics.
5. Estimating abundance and run timing.

# Example data
Load in the example dataset and inspect it. The model requires a date column and a spawner_counts column. Optional observer_efficiency, coverage, and survey_error_group columns can be supplied. Observation efficiency and coverage are used to adjust observed counts, while survey_error_group allows separate observation-dispersion parameters to be estimated for different survey types or quality classes.

In this case, dates are provided as character strings, and observer efficiency and coverage have been provided as percentages. These will be converted into the proper format in the data validation step.


```{r load-data}
data(clemens_sockeye)

summary(clemens_sockeye)
```

# Validate data
The first step is to validate and standardize the input data. This converts the dates into their proper format and converts observer efficiency and coverage to proportions if they were provided as percentages. Validation may generate warnings if missing correction factors are supplied or if some years show potential evidence of multiple run peaks.
```{r validate-data}
clemens_sockeye <- validate_data(clemens_sockeye)

summary(clemens_sockeye)
```

# Survey observations
Now plot the raw survey counts by date in each year.
```{r plot-data, fig.width=8, fig.height=5}
ggplot(clemens_sockeye |> mutate(year = year(date), day = yday(date)),
  aes(x = day, y = spawner_counts, fill = observer_efficiency))+
  geom_point(pch = 21, size = 2) +
  facet_wrap(~year, scales = "free_y")+
  theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(
      labels = scales::label_number(
        scale_cut = scales::cut_short_scale()
      ),
      name = "Spawners observed"
    ) +
    scale_x_continuous(
      labels = function(x)
        format(
          as.Date(x, origin = "2023-12-31"),
          "%b"
        ),
      name = NULL
    )+
  scale_fill_viridis_c(option = "G", direction = -1)
```

# Fit the model
spawnBayes estimates annual run timing and abundance jointly across years.
Information is shared among years through a hierarchical model, improving
estimation in years with sparse survey data.

By default, spawnBayes fits a robust observation model that allows a small proportion of surveys to arise from a higher-variance observation process. This reduces the influence of unusual surveys on abundance and run-timing estimates. The feature can be disabled using outlier_model = FALSE.

The settings used in this vignette are intended to keep computation time
reasonable. For applied analyses, users should always examine the
diagnostic output and increase the number of iterations when recommended. 
Even so, the model may require several minutes to fit.
```{r fit-model, cache=FALSE, message=FALSE, warning=FALSE}
fit <- fit_spawner(clemens_sockeye, iter_warmup = 500, iter_sampling = 500, assumed_residence = 15.5, adapt_delta = 0.99, refresh = 0)
```

# Model diagnostics

The `diagnostics()` function summarizes several indicators of MCMC sampling performance and model convergence. 
```{r diagnostics}
diagnostics(fit)
```
In this example, all diagnostic checks were acceptable. No divergent
transitions were detected, no maximum tree-depth warnings were encountered,
and the largest Rhat value was close to 1. Effective sample sizes for both
bulk and tail regions of the posterior distribution were sufficient for
reliable inference.

When fitting new datasets, users should pay particular attention to:

* Divergences: should be zero whenever possible.
* Tree depth hits: frequent hits may indicate that model settings should be adjusted.
* Rhat: values close to 1 indicate convergence, while values substantially
  greater than 1 suggest poor mixing among chains.
* Effective sample size (ESS): larger values indicate better estimation of
  posterior summaries.

The recommendation column provides a summary of whether additional model
tuning or longer runs are required.

# Prior-posterior comparisons

Prior-posterior comparisons are a useful diagnostic for understanding how much information the data contribute to parameter estimation. When using loosely informative priors, posterior distributions are often expected to be narrower than the priors while remaining largely consistent with them. Prior-posterior comparisons can help identify parameters that are strongly informed by the data and assess whether prior assumptions are reasonable.

```{r prior-post, fig.width=8, fig.height=7}
prior_post_plot(fit)
```

# Estimated run
The fitted curves represent the estimated number of spawners present
through time. Shaded regions show the 66 and 95% credible intervals.
Observed survey counts are shown as points. When the outlier model is enabled, point colour indicates the posterior probability that a survey belongs to the outlier observation component. The open circles show the uncorrected survey counts. The filled circles are corrected for observer efficiency and survey coverage. 

```{r run-curves, fig.width=8, fig.height=7}
plot_run_curve(fit)
```

# Outlier diagnostics  

`spawnBayes` can identify surveys that deviate substantially from the overall run-timing pattern. These observations are not removed from the analysis; instead, the model estimates the probability that they arise from a higher-variance observation process and correspondingly reduces their influence on parameter estimation.

Surveys with elevated outlier probabilities may reflect counting errors, unusual survey conditions, inaccuracies in correction factors, or biological patterns that are not fully captured by the run-timing model. Elevated outlier probabilities do not necessarily indicate erroneous data. Rather, they identify observations that are difficult to reconcile with the fitted observation and run-timing processes.

Users are encouraged to inspect years containing surveys with high outlier probabilities and assess whether the fitted run-timing model is appropriate for estimating spawner abundance.

# Abundance estimates

The abundance summary reports posterior median abundance estimates and credible intervals for each year. When residence time is supplied, traditional trapezoidal area under the curve (TAUC) abundance estimates are also reported, together with the percent difference between the posterior median abundance estimate and the corresponding TAUC estimate.

Two diagnostic indicators are included in the abundance table. The `multimodal` column identifies years with evidence of multiple major run-timing peaks, which may violate the model assumption of a single seasonal peak. The `n_outliers` column reports the number of surveys within each year having posterior outlier probabilities greater than 0.25. These diagnostics should be interpreted as indicators of potential model-data conflict rather than measures of data quality.
```{r abundance}
abund <- abundance(fit)
knitr::kable(abund)
```

The abundance plot shows the median (point) as well as the 66 and 95% posterior credible intervals. The red x shows the estimate based on trapezoidal area under the curve, with an assumed 15.5 day residence (survey life).
```{r abund-plot, fig.width=6, fig.height=4}
plot_abundance(fit)
```
# Timing estimates

Run timing summaries describe the estimated distribution of spawner arrival dates, residence, and exit lag in each year.

```{r timing-plot, fig.width=8, fig.height=8}
plot_timing(fit)
```

# Conclusions

spawnBayes provides a framework for estimating salmon spawner abundance
and run timing from repeated live-count surveys while accounting for
uncertainty in both the observation and process components of the model.

# Citation

The methods implemented in `spawnBayes` are described in:

Thompson, P. L., Akenhead, S. A., and Louie, C. (2026).
*Estimating spawning salmon abundance from visual surveys:
a hierarchical model of arrival and exit dynamics*.
Canadian Journal of Fisheries and Aquatic Sciences, 83, 1-13.
https://doi.org/10.1139/cjfas-2026-0029

Users are encouraged to cite both the package and the accompanying
manuscript when using spawnBayes in scientific publications.